---
title: "Tutorial01 - Asset Returns"
subtitle: "Understanding Financial Returns"
output: 
  html:
    css: ../hw.css
link-citations: true
editor: visual
markdown: 
  wrap: sentence
---

```{r include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  out.width = "80%",
  fig.asp = 0.618,
  fig.width = 10
)
library(tidyverse)
library(moments) # for skewness and kurtosis
library(tseries) # for financial time series functions
```

# Learning Outcomes

By the end of this tutorial, you will be able to:

- Calculate and interpret simple and logarithmic returns
- Understand the differences between various types of returns
- Analyze the statistical properties of financial returns
- Create basic portfolio return calculations
- Visualize and interpret financial time series data

# Prerequisites

This tutorial assumes you have:
- Basic knowledge of R programming
- Understanding of basic statistical concepts
- Familiarity with the tidyverse package
- Access to the provided financial dataset

# Getting Started

## Data Loading

We'll be working with financial market data stored in `fin7028_dat.RData`. This dataset contains:
- Daily stock prices
- Return calculations
- Market index data
- Trading volumes

```{r load-data}
load("fin7028_dat.RData")
```

## Key Concepts

Before we dive into the analysis, let's review some key concepts:

- **Simple Returns**: Percentage change in price
- **Log Returns**: Natural logarithm of price ratios
- **Portfolio Returns**: Weighted combination of asset returns
- **Excess Returns**: Return above a benchmark
- **Volatility**: Measure of return variation


# Additional Resources

- Recommended readings on financial returns
- Links to relevant documentation
- References for statistical concepts used

# Tips for Success

- Always check for missing values in financial data
- Pay attention to time series aspects of the data
- Consider the implications of different return calculations
- Document your analysis clearly
- Interpret results in financial context

::: callout-note
Remember that financial returns often exhibit specific characteristics like:
- Non-normal distributions
- Volatility clustering
- Time-varying relationships
:::